---
title: "Transmission Type and Fuel Efficiency"
author: "Taiki Sakai"
date: "October 24, 2015"
output: 
      pdf_document:
            fig_caption: yes
---

## Summary
In this paper we examine the effect of transmition type (automatic or manual)
on fuel efficiency (miles per gallon). We will try to create a model for
fuel efficiency using any number of variables among weight, horsepower, 
transmission type, and other data recorded for 32 different cars. After considering
a few differnet models, we arrive at a model using horsepower and transmission type
and conclude that an automatic transmission has lower fuel efficiency than a
manual transmission

## Exploratory Analysis
We start our analysis by loading in the dataset and taking a quick look at
the correlation matrix and the pairs plot (Figure 3 in the Appendix). Only a 
subset of the correlation matrix is reproduced, for brevity.
```{r, eval=FALSE, echo=FALSE}
library(datasets)
data(mtcars)
cor(mtcars)
```
```{r,include=FALSE,echo=FALSE}
library(dplyr)
```

```{r, echo=FALSE}
cor(select(mtcars, c(mpg,cyl,disp,hp,wt,am)))
```
From the correlation matrix we can see that horsepower, cylinder count,
displacement, and weight are all highly correlated with each other. This makes sense,
as a higher displacement engine would require more cylinders and should yield a 
higher horsepower. A higher weight vehicle will require more horsepower to
move it effectively. Intuitively, it would make sense if one of (or all of) these variables
would have an effect on fuel efficiency, and we can see thta they all have high correlation
with mpg. Since they are all strongly correlated with each other, it may be possible to use 
just one of them as a proxy for all of the others as they generally measure a similar
quality of a vehicle. The rest of the variables have a lower correlation with mpg,
so these seem like a good place to start our analysis.

We also notice that cyl, disp, and wt all have a relatively high correlation with
am, the other variable we are interested in. Because of this, we will begin our analysis 
using horsepower and transmission type, as they are both correlated with mpg but not
strongly correlated with each other. 

## Model Building
We will start our model at a very basic level. There are clearly factors other
than transmission type that should affect fuel efficiency, so we will not
consider a model that only includes transmission type. Our first model will
use horsepower and transmission type, and we will try to build more complex 
models from there.
```{r,echo=FALSE}
fit1 <- lm(mpg~hp+am, mtcars)
summary(fit1)$coefficients
```
With this model we can see that the coefficients for the intercept, hp, and am
are statistically significant. The R-squared is fairly high at .767, suggesting
the model is a decent fit for our data. The residual plot (Figure 1 in Appendix)
does not seem to have any pattern to it, suggesting that the model might be reasonable.

From here we will try adding other variables from our dataset and use an anova
test to determine if the additional variables contribute to the model. We will include
all of the remaining variables in the dataset, and also an interaction variable
between horsepower and transmission type. For brevity, the results of these are not
shown.

These tests reveal that only one of the additional variables appears to significantly
improve our model. That p-value corresponds to the weight variable. With this new
knowledge we construct a model with horsepower, weight, and transmission type.
```{r, echo=FALSE}
fit2<-lm(mpg~hp+am+wt,mtcars)
summary(fit2)$coefficients
```
One thing to notice is that the additon of the weight variable has increased the standard
error of the coefficient for our transmission variable to the point where it is no
longer statistically significant. This could suggest that there is not a relationship
between fuel efficiency and transmission type. However, we also notice from the residual
plot (Figure 2 in Appendix) that there might be a slight negative or U-shaped pattern in the residuals. This
suggests that the model might not be correct. With this concern in mind, we will
choose to revert to our simpler model using only horsepower and transmission type.

## Results

Our final model is:
```{r, echo=FALSE}
summary(fit1)
```
These coefficients can be interpreted as: for every 1 increase in horsepower, mpg
decreases by .0589. If a transmission is manual (am=1), mpg increases by 5.277. 
Thus we conclude that manual transmissions have higher fuel efficiency, since the 
coefficient 5.277 is positive and is statistically significant. A 95% confidence
interval for the increased mpg in a manual transmission is
`r c(5.277-1.09*qt(.975,29), 5.277+1.09*qt(.975,29))`.

## Appendix

```{r,fig.width=3.5, fig.height=3.5, echo=FALSE, fig.cap='Residuals for Simpler Model'}
plot(fitted(fit1), resid(fit1), xlab='Fitted Values', ylab='Residuals', 
     main='Residual plot for mpg~hp+am')
```

```{r, fig.width=3.5, fig.height=3.5, echo=FALSE, fig.cap='Residuals for Model with Weight'}
plot(fitted(fit2),resid(fit2), xlab='Fitted Values', ylab='Residuals',
     main='Residual plot for mpg~hp+am+wt')
```

```{r, fig.width=6, fig.height=6, echo=FALSE, fig.cap='Pairs Plot for mtcars'}
pairs(mtcars)
```


